\chapter{Background and Related Work}
\label{Chapter2}

\section{Android Malware Detection}

Android malware detection has traditionally relied on static, dynamic, or
hybrid analysis. Static analysis inspects an APK without executing it,
extracting features such as requested permissions, API calls, opcode
sequences, and call graphs; dynamic analysis instead runs the app in a
sandbox and observes runtime behavior. Machine-learning and deep-learning
classifiers built on these features have dominated the field for the past
decade~\cite{xu2019droidevolver,kim2019multimodal,cai2021jowmdroid,liu2024unraveling}.
Two persistent weaknesses motivate the shift toward LLM-based approaches
studied in this thesis. First, such classifiers are prone to concept drift:
performance degrades as malware evolves away from the training
distribution~\cite{barbero2022transcending,pendlebury2019tesseract}.
Second, they typically operate as opaque classifiers, offering little in
the way of human-readable justification for a
verdict~\cite{Li2024-wl}. LLM-based methods, including the LAMD framework
this thesis builds on, aim to address both weaknesses at once: broad
pre-trained knowledge for better generalization under drift, and natural
language reasoning for explainable output.

\section{The LAMD Framework}

LAMD~\cite{Qian2025-ua} is the framework the thesis's first implemented
candidate method builds on. It combines two components. \emph{Key-context
extraction} seeds analysis from suspicious API calls -- data-access APIs
such as \texttt{getPrimaryClip()} and data-transmission (sink) APIs -- and
retains only the code that is control- or data-dependent on those calls via
backward program slicing, so the LLM never has to reason over an entire
decompiled application. \emph{Tier-wise code reasoning} then builds an
understanding of each sliced fragment bottom-up in three passes: a Tier-1
pass summarizes each function's behavior; a Factual Consistency
Verification (FCV) step checks that a summary's claimed dependencies are
actually present in the code (a Data Relationship Coverage score,
thresholded at $\theta = 0.95$ in the original paper) before the summary is
trusted; a Tier-2 pass aggregates function summaries into per-API intent;
and a Tier-3 pass produces the final APK-level verdict together with
supporting evidence.

LAMD was selected as a building block for two reasons directly relevant to
the evaluation goals of this thesis. It targets the same two structural
problems that make naive "feed the whole APK to an LLM" approaches
impractical -- context-window limits and the poor fit between
sequential token-based reasoning and Android's deeply nested call
structure -- and its own evaluation reports better robustness to temporal
distribution drift than the traditional baselines it compares against (see
Chapter~\ref{Chapter3}, RQ2, for the literature's account of this result).
Chapter~\ref{Chapter4} describes how this thesis reimplements LAMD as a
baseline and iterates on it.

\section{Large Language Models for Security Tasks}

Large Language Models have been applied to security tasks more broadly than
malware detection alone. General-purpose LLMs and code-specialized variants
built on the Transformer architecture~\cite{brown2020gpt3} have been shown
to analyze code structure, summarize behavior, and reason about intent
without task-specific training. Existing surveys catalog this broader
application space -- code analysis, vulnerability and threat identification,
and disguised or obfuscated malware
detection~\cite{alkaraki2024exploring,guo2025} -- across security domains
in general, not Android specifically. This thesis narrows that broad
capability to one platform and one task: Android malware detection.
Chapter~\ref{Chapter3} reports a systematic literature review focused
specifically on that intersection.

\section{Related Work}

This section covers related work on Android malware detection prior to and
outside the LLM-based literature; LLM-based approaches specifically are
surveyed through a dedicated systematic literature review in
Chapter~\ref{Chapter3}, which also positions this thesis against existing
secondary studies (Chapter~\ref{Chapter3}, Section~\ref{sec:relwork}).

Traditional Android malware detection is a mature research area. Static
feature-based classifiers such as Drebin and graph-based approaches such as
Malscan established the feature families (permissions, API calls, call-graph
centrality) that later work, including LAMD's own baselines, still compares
against. Two lines of criticism from this literature motivate the
LLM-based direction evaluated in this thesis: robustness to concept drift is
consistently reported as a weak point of purely learning-based
classifiers~\cite{barbero2022transcending,pendlebury2019tesseract}, and
opaque, non-explainable verdicts limit how much a security analyst can
trust or act on a classifier's output~\cite{Li2024-wl}. Both concerns
recur throughout the LLM-based literature reviewed in Chapter~\ref{Chapter3}
and directly shape the evaluation criteria used in Chapter~\ref{Chapter4}.
