\chapter{Experimental Results and Discussion}
\label{Chapter5}

\section{Results}

\textit{[TODO: Once additional candidates from Chapter~\ref{Chapter4} are
implemented, report a final cross-candidate comparison table here using the
metrics defined in Section~\ref{sec:metrics}.]}

The results below are an internal ablation of the LAMD-based candidate
only (baseline reimplementation vs.\ its own optimized versions,
Section~\ref{sec:lamd-candidate}), not yet a comparison against other
candidate methods. They were reconstructed from development logs and prior
analysis sessions rather than freshly re-run for this write-up; the exact
figures should be treated as best-available and re-verified before being
cited as final thesis results.

Table~\ref{tab:cache-ablation} isolates the effect of the semantic cache
alone, on a single 24-CFG APK, comparing no cache against a cold cache
(empty at the start of the run) and a warm cache (already populated by a
prior run on similar code). All three configurations agree on the
MALWARE verdict.

\begin{table}[ht]
\centering
\caption{Effect of the semantic cache on the same 24-CFG APK}
\label{tab:cache-ablation}
\begin{tabular}{@{}lrrr@{}}
\toprule
 & No cache & Cold cache & Warm cache \\
\midrule
LLM calls        & 62      & 30 ($-52\%$)  & \textbf{9 ($-85\%$)} \\
Total tokens      & 146{,}533 & 78{,}749 ($-46\%$) & \textbf{41{,}598 ($-72\%$)} \\
Cache hit rate    & --      & 14/24         & \textbf{24/24} \\
Prediction        & MALWARE & MALWARE       & MALWARE \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:paired-comparison} reports a paired baseline-vs-improved
comparison on 6 fresh AndroZoo malware APKs, run once each through the
unoptimized baseline and the current cache-and-verified-connection design.
One case (\texttt{032EF910\ldots}) is excluded from the reduction column: at
the time of this run it exposed the missing-sink-API gap described in
Section~\ref{sec:lamd-candidate}, so the two pipelines used the same evidence
and reached different verdicts by chance rather than by design; after that
fix, the improved pipeline recovers the correct verdict.

\begin{table}[ht]
\centering
\caption{Paired baseline-vs-improved comparison, 6 malware APKs}
\label{tab:paired-comparison}
\begin{tabular}{@{}lrrrl@{}}
\toprule
APK & Baseline calls & Improved calls & Reduction & Prediction \\
\midrule
008C3C11\ldots & 124 & 54 & 56\% & MALWARE (both) \\
019B34A0\ldots & 97  & 48 & 51\% & MALWARE (both) \\
04C5329F\ldots  & 91  & 28 & 69\% & MALWARE (both) \\
05B7569D\ldots  & 89  & 3  & 97\% & MALWARE (both, post-fix) \\
\bottomrule
\end{tabular}
\end{table}

Across the full 12-APK set, the improved pipeline reached the correct
verdict on 11/12 samples (92\%) after both design fixes described in
Section~\ref{sec:lamd-candidate}, up from 10/12 (83\%) before those fixes.
The one remaining miss was traced to Tier-3's verdict not being fully
deterministic across repeated LLM calls on the same borderline evidence,
rather than to either fixed design issue -- noted here as an open point,
discussed further in Section~\ref{sec:threats-ch5}.

\section{Discussion}

\textit{[TODO: Interpret the results. Compare against the related work
reviewed in Chapter~\ref{Chapter2}. Explain any surprising findings, and
directly answer the research questions stated in Chapter~\ref{Chapter1}.]}

\section{Threats to Validity}\label{sec:threats-ch5}

\textit{[TODO: Discuss limitations of the evaluation, e.g. dataset size/bias,
LLM non-determinism, generalizability of results.]}
